---
title: "01OCT24_viz_and_eda_2"
author: "Stella Koo"
date: "2024-10-01"
output: github_document
---
## Vizualizations_Part2

```{r message = FALSE}
library(tidyverse)

weather_df = 
  rnoaa::meteo_pull_monitors(
    c("USW00094728", "USW00022534", "USS0023B17S"),
    var = c("PRCP", "TMIN", "TMAX"), 
    date_min = "2021-01-01",
    date_max = "2022-12-31") |>
  mutate(
    name = case_match(
      id, 
      "USW00094728" ~ "CentralPark_NY", 
      "USW00022534" ~ "Molokai_HI",
      "USS0023B17S" ~ "Waterhole_WA"),
    tmin = tmin / 10,
    tmax = tmax / 10) |>
  select(name, id, everything())

weather_df
```

### Labels
`labs()` to provide informative axis, plot titles, captions, etc.

* In the following example, `color = "Location"` sets title for the legend that describes color aesthetic.

```{r warning = FALSE}
weather_labels = weather_df |>
  ggplot(aes(x = tmin, y = tmax)) +
  geom_point(aes(color = name), alpha = 0.5) +
  labs(title = "Temperature plot",
    x = "Minimum daily temperature (C)",
    y = "Maxiumum daily temperature (C)",
    color = "Location",
    caption = "Data from the rnoaa package") 

weather_labels
```


### Scales
`scale_x_*` and `scale_y_*`: control over the location and specification of tick marks on the X or Y axis. (where * depends on the type of variable mapped to `x` and `y` -> discrete or continuous)

* `breaks` defines where the tick marks should appear on the axis.
* `labels` defines the text displayed at each tick mark.

```{r warning = FALSE}
weather_scales = weather_labels +
  scale_x_continuous(
    breaks = c(-15, 0, 15),
    labels = c("-15º C", "0", "15"))

weather_scales
```

There are variety of options to use. Need to google to check how to do what I want.

* `limits`: sets the range of values that will be displayed on the x-axis.
* `trans`: transforms the y-axis using the assigned transformation.
* `position`: moves the y-axis to the assigned side of the plot.

```{r warning = FALSE}
weather_labels + 
  scale_x_continuous(
    breaks = c(-15, 0, 15), 
    labels = c("-15ºC", "0", "15"),
    limits = c(-20, 30)) + 
  scale_y_continuous(
    trans = "sqrt", 
    position = "right")
```

There are several ways to produce the desired result:

* `scale_y_sqrt()` added to `ggplot` to transform y-scale.
* `xlim()` controls the plot limits in the X axis.
* `scale_color_hue()` controls the color scale and the name in the plot lengend.
  * `scale_color_hue(h = c(100, 300))`: If you have 3 levels (aka categories), for example, it will divide the hue range into 3 distinct colors spaced between 100 and 300. In this case, from green to purple. 

```{r warning = FALSE}
weather_labels + 
  scale_color_hue(h = c(100, 300))
```

`scale_color_viridis()` from `viridis` package is used to apply a viridis color palette to the plot without having to choose my own. 

* `name` sets the title of lengend to "Location"
* `discrete = TRUE`: tells ggplot2 to use the Viridis color palette for a discrete variable. When set to TRUE, distinct colors are assigned to each level of the categorical variable.

```{r warning = FALSE}
ggp_temp_plot = 
  weather_labels + 
  viridis::scale_color_viridis(
    name = "Location", 
    discrete = TRUE)

ggp_temp_plot
```

In cases data is not discrete, use `viridis::scale_fill_viridis()` function for the fill aesthetic used in histograms, density plots, etc.

## Themes
Used to modify non-data elements of a plot. Control things like background color and location of the the legend. Help with the overall appearance of the plot.

* `legend.position` changes location of the legend.
  * `legend.position = "none"` will remove the legend.

* `theme_gray()`: default theme in `ggplot2`. It features a light gray background with white grid lines.
* `theme_bw()`: a white background and black grid lines.
* `theme_classic()`: classic look for plots.

The ordering of `theme_bw()` and `theme()` matters – `theme()` changes a particular element of the plot’s current “theme”. If you call `theme()` to change the some element and then `theme_bw()`, the changes introduced by `theme()` are overwritten by `theme_bw()`.

```{r warning = FALSE}
ggp_temp_plot + 
  theme_bw() + 
  theme(legend.position = "bottom")
```

#### Learning Assessment
```{r warning = FALSE}
weather_df |>
  ggplot(aes(x = date, y = tmin, color = name)) +
  geom_smooth(se = FALSE) +
  geom_point(aes(size = prcp), alpha = 0.5) +
  labs(title = "Minimum Temperature Per Day",
       x = "Date",
       y = "Minimum Temperature (C)",
       color = "Location",
       size = "Precipitation") +
  viridis::scale_color_viridis(discrete = TRUE) +
  theme_bw() +
  theme(legend.position = "left")
```

### Setting Options
The following code chunk is normally added in the beginning of new file to set figure preferences as global options.

```{r}
knitr::opts_chunk$set(
  fig.width = 6,
  fig.asp = .6,
  out.width = "90%"
)

theme_set(theme_minimal() + theme(legend.position = "bottom"))

options(
  ggplot2.continuous.color = "viridis",
  ggplot2.continuous.fill = "viridis"
)

scale_colour_discrete = scale_colour_viridis_d
scale_fill_discrete = scale_fill_viridis_d
```

### Data argument in `geom_*`
Can split `weather_df` into separate datasets for Central Park and Molokai by using one in the ggplot() call and another in geom_line():

```{r warning = FALSE}
central_park_df =
  weather_df |>
  filter(name == "CentralPark_NY")

molokai_df = 
  weather_df |>
  filter(name == "Molokai_HI")

ggplot(data = molokai_df, aes(x = date, y = tmax, color = name)) +
  geom_point() +
  geom_line(data = central_park_df)
```

### `patchwork` package
Used to show two or three fundamentally different plots in the same graphic.

```{r warning = FALSE}
library(patchwork)

tmax_tmin_p = 
  weather_df |> 
  ggplot(aes(x = tmax, y = tmin, color = name)) + 
  geom_point(alpha = .5) +
  theme(legend.position = "none")

prcp_dens_p = 
  weather_df |> 
  filter(prcp > 0) |> 
  ggplot(aes(x = prcp, fill = name)) + 
  geom_density(alpha = .5) + 
  theme(legend.position = "none")

tmax_date_p = 
  weather_df |> 
  ggplot(aes(x = date, y = tmax, color = name)) + 
  geom_point(alpha = .5) +
  geom_smooth(se = FALSE) + 
  theme(legend.position = "bottom")

(tmax_tmin_p + prcp_dens_p) / tmax_date_p
```

### Data Manipulation

* `forcats::fct_relevel`: Reorders the levels of a factor to a specific order that you define.
* `forcats::fct_reorder`: Reorders the levels of a factor based on the values of another variable.

```{r warning = FALSE}
weather_df |>
  mutate(name = forcats::fct_relevel(name, c("Molokai_HI", "CentralPark_NY", "Waterhole_WA"))) |> 
  ggplot(aes(x = name, y = tmax)) + 
  geom_violin(aes(fill = name), color = "blue", alpha = .5) + 
  theme(legend.position = "bottom")
```

```{r warning = FALSE}
weather_df |>
  mutate(name = forcats::fct_reorder(name, tmax)) |> 
  ggplot(aes(x = name, y = tmax)) + 
  geom_violin(aes(fill = name), color = "blue", alpha = .5) + 
  theme(legend.position = "bottom")
```

It is easier to manipulate the data first and then create the desired plots than finding ways within the plot code:

* Manipulate data to facet panels across the `name` variable, and create separate densities for `tmax` and `tmin` in each panel.

```{r warning = FALSE}
weather_df |>
  select(name, tmax, tmin) |> 
  pivot_longer(
    tmax:tmin,
    names_to = "observation", 
    values_to = "temp") |> 
  ggplot(aes(x = temp, fill = observation)) +
  geom_density(alpha = .5) + 
  facet_grid(~name)
```

#### Examples
```{r message = FALSE, warning = FALSE}
pulse_data = 
  haven::read_sas("./data/public_pulse_data.sas7bdat") |>
  janitor::clean_names() |>
  pivot_longer(
    bdi_score_bl:bdi_score_12m,
    names_to = "visit", 
    names_prefix = "bdi_score_",
    values_to = "bdi") |>
  select(id, visit, everything()) |>
  mutate(
    visit = recode(visit, "bl" = "00m"),
    visit = factor(visit, levels = str_c(c("00", "01", "06", "12"), "m"))) |>
  arrange(id, visit)

ggplot(pulse_data, aes(x = visit, y = bdi)) + 
  geom_boxplot()
```

```{r message = FALSE, warning = FALSE}
pup_data = 
  read_csv("./data/FAS_pups.csv", na = c("NA", ".", "")) |>
  janitor::clean_names() |>
  mutate(
    sex = 
      case_match(
        sex, 
        1 ~ "male", 
        2 ~ "female"))

litter_data = 
  read_csv("./data/FAS_litters.csv", na = c("NA", ".", "")) |>
  janitor::clean_names() |>
  separate(group, into = c("dose", "day_of_tx"), sep = 3)

fas_data = left_join(pup_data, litter_data, by = "litter_number") 

fas_data |> 
  select(sex, dose, day_of_tx, pd_ears:pd_walk) |> 
  pivot_longer(
    pd_ears:pd_walk,
    names_to = "outcome", 
    values_to = "pn_day") |> 
  drop_na() |> 
  mutate(outcome = forcats::fct_reorder(outcome, pn_day, median)) |> 
  ggplot(aes(x = dose, y = pn_day)) + 
  geom_violin() + 
  facet_grid(day_of_tx ~ outcome) # rows ~ columns of the grid
```

