---
title: "26SEP24_viz_and_eda"
author: "Stella Koo"
date: "2024-09-25"
output: github_document
---
```{r message = FALSE}
library(tidyverse)
library(ggridges)

knitr::opts_chunk$set(
  fig.width = 6,
  fig.asp = .6,
  out.width = "90%"
)
```


```{r message = FALSE}
weather_df = 
  rnoaa::meteo_pull_monitors(
    c("USW00094728", "USW00022534", "USS0023B17S"),
    var = c("PRCP", "TMIN", "TMAX"), 
    date_min = "2021-01-01",
    date_max = "2022-12-31") |>
  mutate(
    name = case_match(
      id, 
      "USW00094728" ~ "CentralPark_NY", 
      "USW00022534" ~ "Molokai_HI",
      "USS0023B17S" ~ "Waterhole_WA"),
    tmin = tmin / 10,
    tmax = tmax / 10) |>
  select(name, id, everything())

weather_df
```

## Basic Scatterplot

```{r warning = FALSE}
ggplot(weather_df, aes(x = tmin, y = tmax)) + 
  geom_point()
```

Can also use piping to create same plot:

* `ggp_weather` itself will be a blank plot since it only defines the data and the aesthetic mappings, but geoms have not been added (`+ geom_point()`)

```{r eval = FALSE}
ggp_weather = 
  weather_df |>
  ggplot(aes(x = tmin, y = tmax)) 

ggp_weather + geom_point()
```

## Advanced Scatterplot

* `aes(color = name)` assigns different colors to points based on the unique values of the name variable and will automatically create a legend showing the different categories.
* `alpha` controls the transparency of the points. 0.5 makes the points semi-transparent.
* `geom_smooth(se = FALSE)` adds a smoothed line to the plot. `se = FALSE` turns off the shaded confidence interval (standard error) around the smooth line.
* `facet_grid(. ~ name)` 

```{r warning = FALSE}
ggplot(weather_df, aes(x = tmin, y = tmax)) + 
  geom_point(aes(color = name), alpha = .5) +
  geom_smooth(se = FALSE) +
  facet_grid(. ~ name)
```

In the following code, `x` was set to `date` and y to `tmax`. The scatterplot will show the max temperatures for the months recorded for each location (categorized data).

* `aes(size = prcp)`: size of the points in  plot will be mapped to the values of `prcp` variable. If `prcp` value is large, the point will be bigger, vice versa.

```{r warning = FALSE}
ggplot(weather_df, aes(x = date, y = tmax, color = name)) + 
  geom_point(aes(size = prcp), alpha = .5) +
  geom_smooth(se = FALSE) + 
  facet_grid(. ~ name)
```

#### Learning Assessment

* `method = "lm"` to `geom_smooth` creates a linear regression line.

```{r}
weather_df |> 
  filter(name == "CentralPark_NY") |>
  mutate(tmax_fahr = tmax * 1.8 + 32, 
         tmin_fahr = tmin * 1.8 + 32) |>
  ggplot(aes(x = tmin_fahr, y = tmax_fahr)) + 
  geom_point(alpha = 0.5) +
  geom_smooth(method = "lm", se = FALSE)
```

## Odds and Ends

* `geom_hex()` divides the plot area into regular hexagons, counts the number of observations within each, and maps these counts to the hexagon's fill color.
* `geom_bin2d()` creates a 2D histogram by dividing the plot area into rectangles (bins), counting the number of observations in each, and mapping these counts to the rectangle's fill color.
* `geom_density2d()` adds contours representing density of observations in a 2D plane, effectively visualizing the distribution of data points

```{r warning = FALSE}
ggplot(weather_df, aes(x = tmax, y = tmin)) + 
  geom_hex()
```

#### Learning Assessment
```{r warning = FALSE}
ggplot(weather_df) + geom_point(aes(x = tmax, y = tmin), color = "blue")
ggplot(weather_df) + geom_point(aes(x = tmax, y = tmin, color = "blue"))
```

* In the first attempt, all points will be colored blue, and the color is applied as a fixed aesthetic for all points in the plot.
* In the second attempt, interprets "blue" as a categorical variable rather than a color. Since "blue" is treated as a data value, the plot will assign it to a legend, but it won’t actually color the points blue. Instead, it’s expecting "blue" to be a variable in the dataset.

## Univariate Plots
### geom_histogram()

* `fill = name`: `fill` aesthetic maps to the name variable in the dataset. Different categories of name will be assigned different fill colors in the histogram bars.
* `position = "dodge"`: instead of stacked bars, creates separate bars for each category of name next to each other for easier comparison.
* `binwidth = 2`: This specifies the width of each bin 

```{r warning = FALSE}
ggplot(weather_df, aes(x = tmax, fill = name)) + 
  geom_histogram(position = "dodge", binwidth = 2)
```

### geom_density()

* `adjust = .5`: adjusts the bandwidth (the smoothness) of density estimate. A smaller bandwidth (less than 1) means the density plot will be more detailed and less smooth (more sensitive to small variations in the data).
* `color = "blue"`: specifies that the outline of the density plot will be blue, but the fill colors will still correspond to the name variable.
* `geom_rug()`: This adds small tick marks at the bottom of the plot, representing the individual data points along the x-axis.

```{r warning = FALSE}
ggplot(weather_df, aes(x = tmax, fill = name)) + 
  geom_density(alpha = .4, adjust = .5, color = "blue") +
  geom_rug()
```

### geom_boxplot()
```{r}
ggplot(weather_df, aes(x = name, y = tmax)) + 
  geom_boxplot()
```

### geom_violin()

* `stat_summary()`: This function allows to compute and display summary statistics on the plot.
  * `fun = "median"`: calculates the median of the tmax values for each category in name.
  * `color = "blue"`: median will be represented as a blue line.

```{r warning = FALSE}
ggplot(weather_df, aes(x = name, y = tmax)) + 
  geom_violin(aes(fill = name), alpha = .5) + 
  stat_summary(fun = "median", color = "blue")
```

### geom_density_ridges()

* `scale = .85`: controls the vertical scaling of the density ridges. 
  * By default, the density curves are scaled so that they are spread out evenly. 
  * <1: ridges are compressed vertically, which reduces the overlap between them.
  * >1: would increase the height of the ridges and cause more overlap.

```{r}
ggplot(weather_df, aes(x = tmax, y = name)) + 
  geom_density_ridges(scale = .85)
```

#### Learning Assessment

```{r warning = FALSE}
ggplot(weather_df, aes(x = prcp, fill = name)) +
  geom_histogram(binwidth = 50) +
  xlim(0, 500) +
  ylim(0, 150)
```

```{r warning = FALSE}
ggplot(weather_df, aes(x = name, y = prcp)) + 
  geom_boxplot()
```

```{r warning = FALSE}
ggplot(weather_df, aes(x = prcp, fill = name)) + 
  geom_density(alpha = .4, adjust = .5, color = "blue")
```

```{r warning = FALSE}
ggplot(weather_df, aes(x = name, y = prcp)) + 
  geom_violin(aes(fill = name), alpha = .5) + 
  stat_summary(fun = "median", color = "blue")
```

```{r warning = FALSE}
ggplot(weather_df, aes(x = prcp, y = name)) + 
  geom_density_ridges(scale = .85)
```

## Saving and Embedding Plots
Use `ggsave()` by explicitly creating the figure and exporting. It will guess the file type you prefer and has options for specifying features of the plot.

```{r warning = FALSE}
ggp_weather = 
  ggplot(weather_df, aes(x = tmin, y = tmax)) + 
  geom_point(aes(color = name), alpha = .5) 

ggsave("ggp_weather.pdf", ggp_weather, width = 8, height = 5)
```

Embedding plots to R Markdown: 

* `fig.width` sets the width of the figure in inches
* `fig.asp` stands for aspect ratio. It defines the height of the figure relative to its width (e.g fig.asp = 0.6, height will be 60% of  width).
* `out.width` controls the display size of the figure in the output (e.g 90% of the width of the output container).
* `out.height` specifies the height of the output for figures when rendering document.

```{r}
knitr::opts_chunk$set(
  fig.width = 6,
  fig.asp = .6,
  out.width = "90%"
)
```
